---
title: "Prepare Inputs"
author: "David Ory"
output: 
  html_document:
    theme: cosmo
    toc: yes
---


## Administration

#### Purpose
Prepares inputs for the Population Synthesizer. These procedures replace previous procedures done in SQL. Note the input geographies file use the 2000 Decennial Census PUMA boundaries, as do the five-year 2007 to 2011 PUMS data.  

#### Outputs
1.  Year-specific control tables (maz, taz, and county) written to MySQL database
2.  Year-specific household and person PUMS data written to MySQL database

#### TODO
1. Check the dollar year in the control and read the documentation to make sure the PUMS conversion puts them in year 2010 dollars
2. Re-build the controls from census data, use scripts and apis, check the dollar year for the TAZ controls
3. Make the SOC to occupation relationship a table rather than a series of ifs

## Procedure

#### Overhead
```{r overhead, results = 'hide'}
library(knitr)
suppressMessages(library(dplyr))
library(stringr)
suppressMessages(library(RMySQL))
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#### Paramaters
```{r parameters}
MYSQL_SERVER    = "localhost"
MYSQL_DATABASE  = "pop_syn"
MYSQL_USER_NAME = "root" 

# year of the PUMS data
PUMS_YEAR = "2007_2011"

# input data configured for 2000, 2005, and 2010
# YEAR = "year_2000"
YEAR = "year_2005"
# YEAR = "year_2010"
```

#### Remote file locations
```{r remote-dir}
MYSQL_PASSWORD_FILE <- "C:/Users/dory/Desktop/mysql.csv"

CONTROL_DIR <- "M:/Development/PopulationSynthesizer/PopSyn3/_working/mysql_version/input_data"
PUMS_DIR <- "M:/Development/PopulationSynthesizer/PopSyn3/pums_data/acs2011_5yr/ca"

MAZ_CONTROL_FILE <-    paste(CONTROL_DIR, YEAR, "maz_data.csv", sep = "/")
TAZ_CONTROL_FILE <-    paste(CONTROL_DIR, YEAR, "taz_data.csv", sep = "/")
COUNTY_CONTROL_FILE <- paste(CONTROL_DIR, YEAR, "county_data.csv", sep = "/")
AGE_CONTROL_FILE <-    paste(CONTROL_DIR, YEAR, "persons_by_age.csv", sep = "/")
GEOG_CONTROL_FILE <-   paste(CONTROL_DIR, "geographies.csv", sep = "/")

PUMS_HH_FILE <- paste(PUMS_DIR, "ss11hca.csv", sep = "/")
PUMS_PER_FILE <- paste(PUMS_DIR, "ss11pca.csv", sep = "/")

```

#### Data reads
```{r data-reads}
input_maz <- read.csv(MAZ_CONTROL_FILE, header = TRUE)
input_taz <- read.csv(TAZ_CONTROL_FILE, header = TRUE)
input_county <- read.csv(COUNTY_CONTROL_FILE, header = TRUE)
input_age <- read.csv(AGE_CONTROL_FILE, header = TRUE)
input_geog <- read.csv(GEOG_CONTROL_FILE, header = TRUE)

input_pums_hh <- read.csv(PUMS_HH_FILE, header = TRUE)
input_pums_per <- read.csv(PUMS_PER_FILE, header = TRUE)

```

#### Prepare MAZ controls
```{r prep-maz}
control_totals_maz <- input_maz

# the geographies file reminds us we are using the 2000 Census PUMAs
input_geog <- input_geog %>%
  rename(PUMA = PUMA5CE00)

control_totals_maz <- left_join(control_totals_maz, input_geog, by = c("maz_original"))

```

#### Prepare TAZ controls
```{r prep-taz}
control_totals_taz <- control_totals_maz %>%
  group_by(taz_original) %>%
  summarise(households = sum(households))

# create TAZ crosswalk, checking for PUMA and COUNTY overlaps
taz_geog <- input_geog %>%
  group_by(taz, taz_original, PUMA, COUNTYFP, mtc_county_id, county_name) %>%
  summarise(count = n()) %>%
  select(-count)

check_taz_geog <- taz_geog %>%
  group_by(taz) %>%
  summarise(count = n()) %>%
  filter(count > 1)

control_totals_taz <- left_join(control_totals_taz, taz_geog, by = c("taz_original"))

# bring in income
to_join_taz <- input_taz %>%
  select(taz_original, hh_income_quartile_1 = income_2010_0_30, hh_income_quartile_2 = income_2010_30_60, 
         hh_income_quartile_3 = income_2010_60_100, hh_income_quartile_4 = income_2010_100_up)

control_totals_taz <- left_join(control_totals_taz, to_join_taz, by = c("taz_original"))

remove(taz_geog, to_join_taz)

```

#### Prepare County/Meta controls
```{r prep-county}
control_totals_meta <- input_age %>%
  select(county_full_name, age_category, persons_by_age) %>%
  mutate(county_name = str_replace(county_full_name, " County", "")) %>%
  mutate(age_int = str_replace(age_category, "\\+", "")) %>%
  mutate(age_int = str_replace(age_int, "Ages ", "")) %>%
  mutate(age_int = str_replace(age_int, "Age ", "")) %>%
  mutate(age_int = as.numeric(age_int)) %>%
  mutate(age_00_18 = ifelse(age_int < 19, persons_by_age, 0)) %>%
  mutate(age_19_64 = ifelse(age_int >= 19 & age_int < 65, persons_by_age, 0)) %>%
  mutate(age_65_up = ifelse(age_int >= 65, persons_by_age, 0))

control_totals_meta <- control_totals_meta %>%
  group_by(county_name) %>%
  summarise(age_00_18 = sum(age_00_18), age_19_64 = sum(age_19_64), age_65_up = sum(age_65_up))

# create county crosswalk
county_names <- input_county %>%
  select(county_name_factor = county_name, mtc_county_id) %>%
  mutate(county_name = paste(county_name_factor)) %>%
  select(-county_name_factor)

control_totals_meta <- left_join(control_totals_meta, county_names, by = c("county_name"))

# bring in hh size, hh workers, and occupation
to_join_meta <- input_county %>%
  select(mtc_county_id, hh_size_1, hh_size_2, hh_size_3, hh_size_4_plus, hh_workers_0, hh_workers_1, hh_workers_2, hh_workers_3_plus,
         occupation_management, occupation_professional, occupation_services,
         occupation_retail, occupation_manual, occupation_military)

control_totals_meta <- left_join(control_totals_meta, to_join_meta, by = c("mtc_county_id"))

remove(county_names, to_join_meta)

```

#### Prepare PUMS data
```{r prep-pums}
# get list of relevant pums
relevant_pumas <- unique(input_geog$PUMA)

pums_hh  <- input_pums_hh[input_pums_hh$PUMA %in% relevant_pumas,]
pums_per <- input_pums_per[input_pums_per$PUMA %in% relevant_pumas,]

# remove vacant units and group quarters households
pums_hh <- pums_hh %>%
  filter(!(NP ==0)) %>%
  filter(TYPE <= 1)

# compute number of workers in the household
num_workers <- pums_per %>%
  mutate(workers = ifelse(ESR == 1 | ESR == 2 | ESR == 4 | ESR == 5, 1, 0)) %>%
  group_by(SERIALNO) %>%
  summarise(hh_workers_from_esr = sum(workers))

pums_hh <- left_join(pums_hh, num_workers, by = c("SERIALNO"))

pums_hh <- pums_hh %>%
  mutate(hh_workers_from_esr = ifelse(is.na(hh_workers_from_esr), 0, hh_workers_from_esr))

# put income in constant year dollars (SQL says reported income * rolling reference factor * inflation adjustment)
pums_hh <- pums_hh %>%
  mutate(hh_income_2010 = 999) %>%
  mutate(hh_income_2010 = ifelse(ADJINC == 1102938, HINCP/1.0 * 1.016787 * 1.05156, hh_income_2010)) %>%
  mutate(hh_income_2010 = ifelse(ADJINC == 1063801, HINCP/1.0 * 1.018389 * 1.01265, hh_income_2010)) %>%
  mutate(hh_income_2010 = ifelse(ADJINC == 1048026, HINCP/1.0 * 0.999480 * 1.01651, hh_income_2010)) %>%
  mutate(hh_income_2010 = ifelse(ADJINC == 1039407, HINCP/1.0 * 1.007624 * 1.00000, hh_income_2010)) %>%
  mutate(hh_income_2010 = ifelse(ADJINC == 1018237, HINCP/1.0 * 1.018237 * 0.96942, hh_income_2010))

# give each hh a more manageable unique id
unique_hh_id <- 1:nrow(pums_hh)
pums_hh <- cbind(pums_hh, unique_hh_id)
rownames(pums_hh) <- 1:nrow(pums_hh)

# set the initial person weight as the hh weight and give the person table the hhnum id
hh_weights <- pums_hh %>%
  select(SERIALNO, wgtp = WGTP, unique_hh_id)

pums_per <- left_join(pums_per, hh_weights, by = c("SERIALNO"))

# use ESR to set employment dummy
pums_per <- pums_per %>%
  mutate(employed = ifelse(ESR == 1 | ESR == 2 | ESR == 4 | ESR == 5, 1, 0L))

# extract the occupation code
pums_per <- pums_per %>%
  mutate(soc = str_trim(socp00, side = c("both"))) %>%
  mutate(soc = ifelse(soc == "N.A.//", str_trim(socp10, side = c("both")), soc)) %>%
  mutate(soc = str_sub(soc, start = 1, end = 2)) %>%
  mutate(occupation = 999) %>%
  mutate(occupation = ifelse(soc == 11, 1, occupation)) %>% # TODO check, SQL code says 'management'
  mutate(occupation = ifelse(soc == 13 | soc == 15 | soc == 17 | soc == 19 | soc == 23 | soc == 25 | soc == 29, 2, occupation)) %>% # 'professional'
  mutate(occupation = ifelse(soc == 21 | soc == 27 | soc == 31 | soc == 33 | soc == 39 | soc == 43, 3, occupation)) %>% # 'services'
  mutate(occupation = ifelse(soc == 35 | soc == 41, 4, occupation)) %>% # 'retail'
  mutate(occupation = ifelse(soc == 37 | soc == 45 | soc == 47 | soc == 49 | soc == 51 | soc == 53 , 5, occupation)) %>% # 'manual'
  mutate(occupation = ifelse(soc == 55, 6, occupation)) # 'military'

table(pums_per$occupation)
table(pums_per$soc)

remove(hh_weights, num_workers, relevant_pumas)


```

#### Put data into MySQL database
```{r data-to-mysql}
# get password
mysql_passes <- read.csv(MYSQL_PASSWORD_FILE, header = TRUE)

mysql_passes <- mysql_passes %>%
  filter(user == MYSQL_USER_NAME) %>%
  mutate(pwd = paste(pwd))

# connection
mysql_connection <- dbConnect(MySQL(), user = MYSQL_USER_NAME, password = mysql_passes$pwd, host = MYSQL_SERVER, dbname = MYSQL_DATABASE)

# write the control tables
dbWriteTable(conn = mysql_connection, name = paste('control_totals_maz',  YEAR, sep = '_'), value = as.data.frame(control_totals_maz),  overwrite = TRUE)
dbWriteTable(conn = mysql_connection, name = paste('control_totals_taz',  YEAR, sep = '_'), value = as.data.frame(control_totals_taz),  overwrite = TRUE)
dbWriteTable(conn = mysql_connection, name = paste('control_totals_meta', YEAR, sep = '_'), value = as.data.frame(control_totals_meta), overwrite = TRUE)

# write the household and person tables
dbWriteTable(conn = mysql_connection, name = paste('household_table', PUMS_YEAR, sep = '_'), value = as.data.frame(pums_hh),  overwrite = TRUE)
dbWriteTable(conn = mysql_connection, name = paste('person_table', PUMS_YEAR, sep = '_'),    value = as.data.frame(pums_per), overwrite = TRUE)

dbDisconnect(mysql_connection)
```

