---
title: "Step 03 Controls to Database"
author: "David Ory"
output: 
  html_document:
    theme: cosmo
    toc: yes
---


## Administration

#### Purpose
Prepares database inputs for the Population Synthesizer. These procedures replace previous procedures done in SQL. Note the input geographies file use the 2000 Decennial Census PUMA boundaries, as do the five-year 2007 to 2011 PUMS data.  

#### Outputs
1.  Year-specific control tables (maz, taz, and county) written to MySQL database

#### TODO
1.  

## Procedure

#### Overhead
```{r overhead, results = 'hide'}
library(knitr)
suppressMessages(library(dplyr))
library(stringr)
suppressMessages(library(RMySQL))
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#### Paramaters
```{r parameters}
MYSQL_SERVER    = "localhost"
MYSQL_DATABASE  = "pop_syn"
MYSQL_USER_NAME = "root" 

# input data configured for 2000, 2005, and 2010
# YEAR = "year_2000"
# YEAR = "year_2005"
YEAR = "year_2010"

```

#### Remote file locations
```{r remote-dir}
MYSQL_PASSWORD_FILE <- "C:/Users/dory/Desktop/mysql.csv"

CONTROL_DIR <- "M:/Development/PopulationSynthesizer/PopSyn3/_working/mysql_version/input_data"

MAZ_CONTROL_FILE <-    paste(CONTROL_DIR, YEAR, "maz_data.csv", sep = "/")
TAZ_CONTROL_FILE <-    paste(CONTROL_DIR, YEAR, "taz_data.csv", sep = "/")
COUNTY_CONTROL_FILE <- paste(CONTROL_DIR, YEAR, "county_controls.csv", sep = "/")
GEOG_CONTROL_FILE <-   paste(CONTROL_DIR, "geographies.csv", sep = "/")

```

#### Data reads
```{r data-reads}
input_maz <- read.csv(MAZ_CONTROL_FILE, header = TRUE)
input_taz <- read.csv(TAZ_CONTROL_FILE, header = TRUE)
input_county <- read.csv(COUNTY_CONTROL_FILE, header = TRUE, stringsAsFactors = FALSE)
input_geog <- read.csv(GEOG_CONTROL_FILE, header = TRUE)

```

#### Prepare MAZ controls
```{r prep-maz}
control_totals_maz <- input_maz

# the geographies file reminds us we are using the 2000 Census PUMAs
input_geog <- input_geog %>%
  rename(PUMA = PUMA5CE00)

control_totals_maz <- left_join(control_totals_maz, input_geog, by = c("maz_original"))

```

#### Prepare TAZ controls
```{r prep-taz}
control_totals_taz <- control_totals_maz %>%
  group_by(taz_original) %>%
  summarise(households = sum(households))

# create TAZ crosswalk, checking for PUMA and COUNTY overlaps
taz_geog <- input_geog %>%
  group_by(taz, taz_original, PUMA, COUNTYFP, mtc_county_id, county_name) %>%
  summarise(count = n()) %>%
  select(-count)

check_taz_geog <- taz_geog %>%
  group_by(taz) %>%
  summarise(count = n()) %>%
  filter(count > 1)

control_totals_taz <- left_join(control_totals_taz, taz_geog, by = c("taz_original"))

# bring in income
to_join_taz <- input_taz %>%
  select(taz_original, hh_income_quartile_1 = income_2010_0_30, hh_income_quartile_2 = income_2010_30_60, 
         hh_income_quartile_3 = income_2010_60_100, hh_income_quartile_4 = income_2010_100_up)

control_totals_taz <- left_join(control_totals_taz, to_join_taz, by = c("taz_original"))

remove(taz_geog, to_join_taz)

```

#### Prepare County/Meta controls
```{r prep-county}
# bring in hh size, hh workers, and occupation
control_totals_meta <- input_county

# create county crosswalk
county_names <- input_county %>%
  select(county_name_factor = county_name, mtc_county_id) %>%
  mutate(county_name = paste(county_name_factor)) %>%
  select(-county_name_factor)

control_totals_meta <- left_join(control_totals_meta, county_names, by = c("county_name", "mtc_county_id"))

remove(county_names)

```

#### Put data into MySQL database
```{r data-to-mysql}
# get password
mysql_passes <- read.csv(MYSQL_PASSWORD_FILE, header = TRUE)

mysql_passes <- mysql_passes %>%
  filter(user == MYSQL_USER_NAME) %>%
  mutate(pwd = paste(pwd))

# connection
mysql_connection <- dbConnect(MySQL(), user = MYSQL_USER_NAME, password = mysql_passes$pwd, host = MYSQL_SERVER, dbname = MYSQL_DATABASE)

# write the control tables
dbWriteTable(conn = mysql_connection, name = paste('control_totals_maz',  YEAR, sep = '_'), value = as.data.frame(control_totals_maz),  overwrite = TRUE)
dbWriteTable(conn = mysql_connection, name = paste('control_totals_taz',  YEAR, sep = '_'), value = as.data.frame(control_totals_taz),  overwrite = TRUE)
dbWriteTable(conn = mysql_connection, name = paste('control_totals_meta', YEAR, sep = '_'), value = as.data.frame(control_totals_meta), overwrite = TRUE)

dbDisconnect(mysql_connection)
```

