---
title: "Step 02 Build Controls"
author: "David Ory"
output: 
  html_document:
    theme: cosmo
    toc: yes
---


## Administration

#### Purpose
Builds the control tables for the Population Synthesizer from Census data. The controls are then put into a MySQL database using `Prepare Inputs.Rmd`. The Java Population Synthesizer software then reads/writes to MySQL (see `go_go_popsyn.bat`). [Census API docs](http://api.census.gov/data.html). 

#### Outputs
1.  Year-specific control tables in CSV format

#### TODO
1. everything, early days
2. build all county controls here and try to do TAZ with blockgroup to TAZ correspondence

## Procedure

#### Overhead
```{r overhead, results = 'hide'}
library(knitr)

# devtools::install_github("hrecht/censusapi")
suppressMessages(library(censusapi))
suppressMessages(library(RMySQL))
library(stringr)
library(reshape2)
suppressMessages(library(dplyr))
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#### Paramaters
```{r parameters}
# year of the PUMS data
# PUMS_YEAR = "2007_2011"
# PUMS_YEAR = "2000"

# input data configured for 2000, 2005, and 2010
# YEAR = "year_2000"
# YEAR = "year_2005"
# YEAR = "year_2010"

STATE_FIPS = "06"
COUNTY_FIPS_STRING = "01,13,41,55,75,81,85,95,97"

county_codes <- data.frame(county_name = c("Alameda", "Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
                           county = c("001","013","041","055","075","081","085","095","097"), 
                           mtc_county_id = c(4,5,9,7,1,2,3,6,8),
                           stringsAsFactors = FALSE)

occupation_df = data.frame(occupation = c(1,2,3,4,5,6), 
                           occupation_category = c("Management", "Professional", "Services", "Retail", "Manual", "Military"),
                           stringsAsFactors = FALSE)

MYSQL_SERVER    = "localhost"
MYSQL_DATABASE  = "pop_syn"
MYSQL_USER_NAME = "root" 
```

#### Remote file locations
```{r remote-dir}
CENSUS_API_KEY_FILE <- "C:/Users/dory/Desktop/census_api_key.csv"

MYSQL_PASSWORD_FILE <- "C:/Users/dory/Desktop/mysql.csv"

CONTROL_DIR <- "M:/Development/PopulationSynthesizer/PopSyn3/_working/mysql_version/input_data"

# TODO add year 2000 pums to vector
# TODO try reading the 2000 pums from the current pop syn dir into MySQL in Prepare Inputs.Rmd
PUMS_DIR <- c("M:/Development/PopulationSynthesizer/PopSyn3/pums_data/acs2011_5yr/ca")
PUMS_HH_FILE <- c(paste(PUMS_DIR, "ss11hca.csv", sep = "/"))
PUMS_PER_FILE <- c(paste(PUMS_DIR, "ss11pca.csv", sep = "/"))

CONTROL_DIR <- "M:/Development/PopulationSynthesizer/PopSyn3/_working/mysql_version/input_data"
PUMS_DIR <- "M:/Development/PopulationSynthesizer/PopSyn3/pums_data/acs2011_5yr/ca"

GEOG_CONTROL_FILE <-   paste(CONTROL_DIR, "geographies.csv", sep = "/")

```

#### Get the Census API key
```{r get-key}
api_key <- read.csv(CENSUS_API_KEY_FILE, header = TRUE)
api_key <- api_key %>%
  mutate(key = paste(key))
```

#### Build the PUMA to County cross-walk
```{r puma-county}
input_geog <- read.csv(GEOG_CONTROL_FILE, header = TRUE, stringsAsFactors = FALSE)

puma_to_county <- input_geog %>%
  select(PUMA = PUMA5CE00, county_name) %>%
  group_by(PUMA, county_name) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  select(-count)


remove(input_geog)
```

# Data reads
```{r data-reads}
# get password
mysql_passes <- read.csv(MYSQL_PASSWORD_FILE, header = TRUE)

mysql_passes <- mysql_passes %>%
  filter(user == MYSQL_USER_NAME) %>%
  mutate(pwd = paste(pwd))

# connection
mysql_connection <- dbConnect(MySQL(), user = MYSQL_USER_NAME, password = mysql_passes$pwd, host = MYSQL_SERVER, dbname = MYSQL_DATABASE)

# read the synthetic population tables
pums_person_0711 <- dbReadTable(conn = mysql_connection, name = paste('person_table', "2007_2011", sep = '_'))
pums_person_00   <- dbReadTable(conn = mysql_connection, name = paste('person_table', "2000", sep = '_'))

dbDisconnect(mysql_connection)

```

#### County Control: Households by size
```{r county-household-size}
# 2000: use decennial, SF1
raw_census <- getCensus(name = "sf1", 
                        vintage = 2000, 
                        key = api_key$key, 
                        vars=c("H013001", "H013002", "H013003", "H013004", "H013005", "H013006", "H013007", "H013008"), 
                        region = paste("county:", COUNTY_FIPS_STRING, sep = ""),
                        regionin = paste("state:", STATE_FIPS, sep = ""))

working <- left_join(raw_census, county_codes, by = c("county"))

household_size_df <- working %>%
  rename(hh_size_1 = H013002, hh_size_2 = H013003, hh_size_3 = H013004) %>%
  mutate(hh_size_4_plus = H013005 + H013006 + H013007 + H013008) %>%
  mutate(year = 2000) %>%
  select(year, county_name, mtc_county_id, hh_size_1, hh_size_2, hh_size_3, hh_size_4_plus)

# 2005
# the 2005 to 2009 estimate is availabe via API, the 2005 is not, but it is available via American Fact Finder: http://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_05_EST_B11016&prodType=table
# for now, enter manually
county_name = c("Alameda", "Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma")
hh_size_1      = c(138023,  86591, 30247, 13619, 135391, 64287, 138519, 27870, 48213)
hh_size_2      = c(158682, 117955, 37125, 16347, 102717, 80142, 167536, 40804, 59038)
hh_size_3      = c( 84858,  62441, 14124,  6699,  37105, 43252, 101650, 25748, 28744)
hh_size_4_plus = c(139817,  95115, 20897, 11537,  47186, 66546, 172425, 40202, 41217)

working <- data.frame(county_name, hh_size_1, hh_size_2, hh_size_3, hh_size_4_plus, stringsAsFactors = FALSE)

working <- left_join(working, county_codes, by = c("county_name"))

working <- working %>%
  mutate(year = 2005) %>%
  select(-county)

household_size_df <- rbind(household_size_df, working)


# 2010: use decennial, SF1 (note that variable names are different than for 2000)
raw_census <- getCensus(name = "sf1", 
                        vintage = 2010, 
                        key = api_key$key,
                        vars=c("H0130001", "H0130002", "H0130003", "H0130004", "H0130005", "H0130006", "H0130007", "H0130008"), 
                        region = paste("county:", COUNTY_FIPS_STRING, sep = ""),
                        regionin = paste("state:", STATE_FIPS, sep = ""))

working <- left_join(raw_census, county_codes, by = c("county"))

working <- working %>%
  rename(hh_size_1 = H0130002, hh_size_2 = H0130003, hh_size_3 = H0130004) %>%
  mutate(hh_size_4_plus = H0130005 + H0130006 + H0130007 + H0130008) %>%
  mutate(year = 2010) %>%
  select(year, county_name, mtc_county_id, hh_size_1, hh_size_2, hh_size_3, hh_size_4_plus)

household_size_df <- rbind(household_size_df, working)

remove(working, raw_census)
```

#### County Control: Household and Group Quarters Population
```{r county-population}
# 2000: use decennial, SF1
raw_census <- getCensus(name = "sf1", 
                        vintage = 2000, 
                        key = api_key$key, 
                        vars=c("H010001", "P037006"), 
                        region = paste("county:", COUNTY_FIPS_STRING, sep = ""),
                        regionin = paste("state:", STATE_FIPS, sep = ""))

working <- left_join(raw_census, county_codes, by = c("county"))

hh_gq_pop_df <- working %>%
  rename(hh_population = H010001, non_inst_gq_population = P037006) %>%
  mutate(year = 2000) %>%
  select(year, county_name, mtc_county_id, hh_population, non_inst_gq_population)

# 2010: use decennial, SF1 (note that variable names are different than for 2000)
raw_census <- getCensus(name = "sf1", 
                        vintage = 2010, 
                        key = api_key$key, 
                        vars=c("H0100001", "P0420007"), 
                        region = paste("county:", COUNTY_FIPS_STRING, sep = ""),
                        regionin = paste("state:", STATE_FIPS, sep = ""))

working <- left_join(raw_census, county_codes, by = c("county"))

working <- working %>%
  rename(hh_population = H0100001, non_inst_gq_population = P0420007) %>%
  mutate(year = 2010) %>%
  select(year, county_name, mtc_county_id, hh_population, non_inst_gq_population)

hh_gq_pop_df <- rbind(hh_gq_pop_df, working)

# 2005: use ACS, which is not available via API.
# http://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_05_EST_B25008&prodType=table
# enter hh population manually and use the average of 2000 and 2010 for group quarters
county_name = c("Alameda", "Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma")
hh_population = c(1421308, 1006486, 235609, 127445, 719077, 689271, 1669890, 395426, 453850)
working <- data.frame(county_name, hh_population, stringsAsFactors = FALSE)

gq_2000_df <- hh_gq_pop_df %>%
  filter(year == 2000) %>%
  select(county_name, gq_2000 = non_inst_gq_population)

gq_2010_df <- hh_gq_pop_df %>%
  filter(year == 2010) %>%
  select(county_name, gq_2010 = non_inst_gq_population)

gq_2005_df <- left_join(gq_2000_df, gq_2010_df, by = c("county_name"))

gq_2005_df <- gq_2005_df %>%
  mutate(non_inst_gq_population = round((gq_2000 + gq_2010)/2.0)) %>%
  select(-gq_2000, -gq_2010)

working <- left_join(working, gq_2005_df, by = c("county_name"))
working <- left_join(working, county_codes, by = c("county_name"))

working <- working %>%
  mutate(year = 2005) %>%
  select(-county)

hh_gq_pop_df <- rbind(hh_gq_pop_df, working)

remove(gq_2000_df, gq_2005_df, gq_2010_df, working, raw_census)

```

#### County Control: Occupation
```{r county-occupation}
# 2000: use PUMS
relevant_pumas <- unique(puma_to_county$PUMA)

pums_per <- pums_person_00[pums_person_00$PUMA %in% relevant_pumas,]

# TODO fix the NA removal -- why are weights NA?
working <- pums_per %>%
  select(PUMA, occupation, wgtp) %>%
  filter(!(is.na(wgtp))) %>%
  group_by(PUMA, occupation) %>%
  summarise(persons = sum(wgtp)) %>%
  ungroup() %>%
  filter(!(is.na(occupation)))

sum(working$persons)

working <- left_join(working, occupation_df, by = c("occupation"))
working <- left_join(working, puma_to_county, by = c("PUMA"))

working_sum <- working %>%
  mutate(occupation_category = paste("occupation", str_to_lower(occupation_category), sep = "_")) %>%
  group_by(county_name, occupation_category) %>%
  summarise(persons = sum(persons))

persons_occupation_df <- dcast(working_sum, county_name ~ occupation_category, sum, value.var = c("persons"))

persons_occupation_df <- persons_occupation_df %>%
  mutate(year = 2000)

# 2010: use 2007 to 2011 PUMS
pums_per <- pums_person_0711[pums_person_0711$PUMA %in% relevant_pumas,]

sum(pums_per$PWGTP)

working <- pums_per %>%
  select(PUMA, occupation, PWGTP) %>%
  group_by(PUMA, occupation) %>%
  summarise(persons = sum(PWGTP)) %>%
  ungroup() %>%
  filter(!(is.na(occupation)))

sum(working$persons)

working <- left_join(working, occupation_df, by = c("occupation"))
working <- left_join(working, puma_to_county, by = c("PUMA"))

working_sum <- working %>%
  mutate(occupation_category = paste("occupation", str_to_lower(occupation_category), sep = "_")) %>%
  group_by(county_name, occupation_category) %>%
  summarise(persons = sum(persons))

working_reshape <- dcast(working_sum, county_name ~ occupation_category, sum, value.var = c("persons"))

working_reshape <- working_reshape %>%
  mutate(year = 2010)

persons_occupation_df <- rbind(persons_occupation_df, working_reshape)

# 2005: use the average of the two PUMS estimates
occ_2000 <- persons_occupation_df %>%
  filter(year == 2000) %>%
  select(-year)
occ_2000 <- melt(occ_2000, id = c("county_name"), variable.name = "occupation_category", value.name = "persons_2000", factorsAsStrings = FALSE)

occ_2010 <- persons_occupation_df %>%
  filter(year == 2010) %>%
  select(-year)
occ_2010 <- melt(occ_2010, id = c("county_name"), variable.name = "occupation_category", value.name = "persons_2010", factorsAsStrings = FALSE)

occ_2005 <- left_join(occ_2000, occ_2010, by = c("county_name", "occupation_category"))

occ_2005 <- occ_2005 %>%
  mutate(persons = round((persons_2000 + persons_2010) / 2.0)) %>%
  select(-persons_2000, -persons_2010)

working_reshape <- dcast(occ_2005, county_name ~ occupation_category, sum, value.var = c("persons"))

working_reshape <- working_reshape %>%
  mutate(year = 2005)

persons_occupation_df <- rbind(persons_occupation_df, working_reshape)

remove(working, working_sum, occ_2000, occ_2005, occ_2010)

```

#### County Control: Age
```{r county-age}
# 2000: use decennial, SF1
raw_census <- getCensus(name = "sf1", 
                        vintage = 2000, 
                        key = api_key$key, 
                        vars=c("P012003", "P012004", "P012005", "P012006", "P012007", "P012008", "P012009", "P012010",
                               "P012011", "P012012", "P012013", "P012014", "P012015", "P012016", "P012017", "P012018",
                               "P012019", "P012020", "P012021", "P012022", "P012023", "P012024", "P012025",
                               "P012027", "P012028", "P012029", "P012030", "P012031", "P012032", "P012033", "P012034", 
                               "P012035", "P012036", "P012037", "P012038", "P012039", "P012040", "P012041", "P012042", 
                               "P012043", "P012044", "P012045", "P012046", "P012047", "P012048", "P012049"), 
                        region = paste("county:", COUNTY_FIPS_STRING, sep = ""),
                        regionin = paste("state:", STATE_FIPS, sep = ""))

working <- left_join(raw_census, county_codes, by = c("county"))

age_df <- working %>%
  mutate(age_00_05 = P012003 + P012027) %>%
  mutate(age_05_09 = P012004 + P012028) %>%
  mutate(age_10_14 = P012005 + P012029) %>%
  mutate(age_15_17 = P012006 + P012030) %>%
  mutate(age_18_19 = P012007 + P012031) %>%
  mutate(age_20_20 = P012008 + P012032) %>%
  mutate(age_21_21 = P012009 + P012033) %>%
  mutate(age_22_24 = P012010 + P012034) %>%
  mutate(age_25_29 = P012011 + P012035) %>%
  mutate(age_30_34 = P012012 + P012036) %>%
  mutate(age_35_39 = P012013 + P012037) %>%
  mutate(age_40_44 = P012014 + P012038) %>%
  mutate(age_45_49 = P012015 + P012039) %>%
  mutate(age_50_54 = P012016 + P012040) %>%
  mutate(age_55_59 = P012017 + P012041) %>%
  mutate(age_60_61 = P012018 + P012042) %>%
  mutate(age_62_64 = P012019 + P012043) %>%
  mutate(age_65_66 = P012020 + P012044) %>%
  mutate(age_67_69 = P012021 + P012045) %>%
  mutate(age_70_74 = P012022 + P012046) %>%
  mutate(age_75_79 = P012023 + P012047) %>%
  mutate(age_80_84 = P012024 + P012048) %>%
  mutate(age_85_up = P012025 + P012049) %>%
  mutate(age_00_18 = age_00_05 + age_05_09 + age_10_14 + age_15_17 + round(0.5 * age_18_19)) %>%
  mutate(age_19_64 = round(0.5 * age_18_19) + age_20_20 + age_21_21 + age_22_24 + age_25_29 + age_30_34 + age_35_39 + age_40_44 + age_45_49 +
           age_50_54 + age_55_59 + age_60_61 + age_62_64) %>%
  mutate(age_65_up = age_65_66 + age_67_69 + age_70_74 + age_75_79 + age_80_84 + age_85_up) %>%
  mutate(age_00_14 = age_00_05 + age_05_09 + age_10_14) %>%
  mutate(age_15_19 = age_15_17 + age_18_19) %>%
  mutate(age_20_44 = age_20_20 + age_21_21 + age_22_24 + age_25_29 + age_30_34 + age_35_39 + age_40_44) %>%
  mutate(age_45_64 = age_45_49 + age_50_54 + age_55_59 + age_60_61 + age_62_64) %>%
  mutate(year = 2000) %>%
  select(year, county_name, mtc_county_id, age_00_18, age_19_64, age_65_up, age_00_14, age_15_19, age_20_44, age_45_64)

# START HERE: 2010 (SF1), then 2005 (?)

```

