---
title: "Consume Outputs"
author: "David Ory"
output: 
  html_document:
    theme: cosmo
    toc: yes
---


## Administration

#### Purpose
Consumes outputs from the Population Synthesizer. These procedures replace previous procedures done in SQL. A PUMS data dictionary is available [here](http://www2.census.gov/programs-surveys/acs/tech_docs/pums/data_dict/PUMS_Data_Dictionary_2007-2011.pdf).   

#### Outputs
1.  Database to be used with Tableau (see `validation/Validation Test.twb` saved as `Year 20XX Validation Test YY.twb`)

#### TODO
1. Add non-control validation variables
2. Non-control check brainstorm: student status (county),

## Procedure

#### Overhead
```{r overhead, results = 'hide'}
library(knitr)
suppressMessages(library(dplyr))
library(stringr)
suppressMessages(library(RMySQL))
suppressMessages(library(hydroGOF))
library(reshape2)
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#### Paramaters
```{r parameters}
MYSQL_SERVER    = "localhost"
MYSQL_DATABASE  = "pop_syn"
MYSQL_USER_NAME = "root" 

# simulation year
# YEAR = "year_2000"
YEAR = "year_2005"
# YEAR = "year_2010"

# year of pums source
PUMS_YEAR = "2007_2011"

HH_INCOME_QUARTILE_THRESH_1 = 30000
HH_INCOME_QUARTILE_THRESH_2 = 60000
HH_INCOME_QUARTILE_THRESH_3 = 100000
```

#### Remote file locations
```{r remote-dir}
MYSQL_PASSWORD_FILE <- "C:/Users/dory/Desktop/mysql.csv"
F_OUTPUT <- paste("~/GitHub/popsyn3/application/validation/", YEAR, "_test_04.csv", sep = "")
```

#### Data reads
```{r data-reads}
# get password
mysql_passes <- read.csv(MYSQL_PASSWORD_FILE, header = TRUE)

mysql_passes <- mysql_passes %>%
  filter(user == MYSQL_USER_NAME) %>%
  mutate(pwd = paste(pwd))

# connection
mysql_connection <- dbConnect(MySQL(), user = MYSQL_USER_NAME, password = mysql_passes$pwd, host = MYSQL_SERVER, dbname = MYSQL_DATABASE)

# read the synthetic population tables
household_df <- dbReadTable(conn = mysql_connection, name = paste('synpop_hh', YEAR, sep = '_'))
person_df <- dbReadTable(conn = mysql_connection, name = paste('synpop_person', YEAR, sep = '_'))

# read in the controls for validation
maz_controls <-    dbReadTable(conn = mysql_connection, name = paste('control_totals_maz', YEAR, sep = '_'))
taz_controls <-    dbReadTable(conn = mysql_connection, name = paste('control_totals_taz', YEAR, sep = '_'))
county_controls <- dbReadTable(conn = mysql_connection, name = paste('control_totals_meta', YEAR, sep = '_'))

# read in the relevant PUMS data to build cross-validation summaries
pums_household <- dbReadTable(conn = mysql_connection, name = paste('household_table', PUMS_YEAR, sep = '_'))
pums_person    <- dbReadTable(conn = mysql_connection, name = paste('person_table', PUMS_YEAR, sep = '_'))

dbDisconnect(mysql_connection)

```

#### MAZ for Controlled Measures
```{r maz-controls}
# maz controls: households
maz_validation <- maz_controls %>%
  select(maz, control_households = households) %>%
  mutate(measure = "households")

maz_synthezised <- household_df %>%
  group_by(maz) %>%
  summarise(synthesized_households = sum(finalweight))

maz_validation <- left_join(maz_validation, maz_synthezised, by = c("maz"))

# replace NAs with zeros
maz_validation <- maz_validation %>%
  mutate(synthesized_households = ifelse(is.na(synthesized_households), 0L, synthesized_households))

# standardize variables to combine across geographies
maz_validation <- maz_validation %>%
  rename(geography_index = maz) %>%
  mutate(geography = "maz") %>%
  mutate(control_persons = NA) %>%
  mutate(synthesized_persons = NA) %>%
  mutate(control_category = "controlled") %>%
  mutate(measure_category = "Households")

remove(maz_synthezised)

```

#### TAZ for Controlled Measures
```{r taz-controls}
taz_validation <- taz_controls %>%
  select(taz, hh_income_quartile_1, hh_income_quartile_2, hh_income_quartile_3, hh_income_quartile_4)

taz_validation <- melt(taz_validation, id = c("taz"), variable.name = "measure", value.name = "control_households")

taz_validation <- taz_validation %>%
  mutate(measure = paste(measure))

taz_synthesized <- household_df %>%
  mutate(income_quartile = 0L) %>%
  mutate(income_quartile = ifelse(hh_income_2010 < HH_INCOME_QUARTILE_THRESH_1, 1L, income_quartile)) %>%
  mutate(income_quartile = ifelse(hh_income_2010 >= HH_INCOME_QUARTILE_THRESH_1 & hh_income_2010 < HH_INCOME_QUARTILE_THRESH_2, 2L, income_quartile)) %>%
  mutate(income_quartile = ifelse(hh_income_2010 >= HH_INCOME_QUARTILE_THRESH_2 & hh_income_2010 < HH_INCOME_QUARTILE_THRESH_3, 3L, income_quartile)) %>%
  mutate(income_quartile = ifelse(hh_income_2010 >= HH_INCOME_QUARTILE_THRESH_3, 4L, income_quartile)) %>%
  group_by(taz, income_quartile) %>%
  summarise(synthesized_households = sum(finalweight)) %>%
  mutate(measure = paste("hh_income_quartile", income_quartile, sep = "_")) %>%
  select(-income_quartile)

taz_validation <- left_join(taz_validation, taz_synthesized, by = c("taz", "measure"))  

# replace NAs with zeros
taz_validation <- taz_validation %>%
  mutate(synthesized_households = ifelse(is.na(synthesized_households), 0L, synthesized_households))

# standardize variables to combine across geographies
taz_validation <- taz_validation %>%
  rename(geography_index = taz) %>%
  mutate(geography = "taz") %>%
  mutate(control_persons = NA) %>%
  mutate(synthesized_persons = NA) %>%
  mutate(control_category = "controlled") %>%
  mutate(measure_category = "Households by Income")

remove(taz_synthesized)

```

#### County for Controlled Measures
```{r county-controls}
# county controls: household size, household workers, age category, occupation category
county_validation <- melt(county_controls, id = c("county_name", "mtc_county_id"), variable.name = "measure", value.name = "households_or_persons")

county_validation <- county_validation %>%
  mutate(measure = paste(measure))

# household size
county_synthesized <- household_df %>%
  group_by(mtc_county_id, np) %>%
  summarise(synthesized_households = sum(finalweight)) %>%
  mutate(measure = "ERROR") %>%
  mutate(measure = ifelse(np == 1, "hh_size_1", measure)) %>%
  mutate(measure = ifelse(np == 2, "hh_size_2", measure)) %>%
  mutate(measure = ifelse(np == 3, "hh_size_3", measure)) %>%
  mutate(measure = ifelse(np > 3,  "hh_size_4_plus", measure)) %>%
  group_by(mtc_county_id, measure) %>%
  summarise(synthesized_households = sum(synthesized_households)) %>%
  mutate(measure_category_household = "Households by Number of People")

# household workers
working <- household_df %>%
  group_by(mtc_county_id, hh_workers_from_esr) %>%
  summarise(synthesized_households = sum(finalweight)) %>%
  mutate(measure = "ERROR") %>%
  mutate(measure = ifelse(hh_workers_from_esr == 0, "hh_workers_0", measure)) %>%
  mutate(measure = ifelse(hh_workers_from_esr == 1, "hh_workers_1", measure)) %>%
  mutate(measure = ifelse(hh_workers_from_esr == 2, "hh_workers_2", measure)) %>%
  mutate(measure = ifelse(hh_workers_from_esr > 2,  "hh_workers_3_plus", measure)) %>%
  group_by(mtc_county_id, measure) %>%
  summarise(synthesized_households = sum(synthesized_households)) %>%
  mutate(measure_category_household = "Households by Number of Workers")

county_synthesized <- rbind(county_synthesized, working)

# done with household level --> join
county_validation <- left_join(county_validation, county_synthesized, by = c("mtc_county_id", "measure"))

# persons by age category
county_synthesized <- person_df %>%
  group_by(mtc_county_id, agep) %>%
  summarise(synthesized_persons = sum(finalweight)) %>%
  mutate(measure = "ERROR") %>%
  mutate(measure = ifelse(agep <= 18, "age_00_18", measure)) %>%
  mutate(measure = ifelse(agep >= 19 & agep < 65, "age_19_64", measure)) %>%
  mutate(measure = ifelse(agep >= 65, "age_65_up", measure)) %>%
  group_by(mtc_county_id, measure) %>%
  summarise(synthesized_persons = sum(synthesized_persons)) %>%
  mutate(measure_category_person = "Persons by Age Category")

# persons by occupation
working <- person_df %>%
  group_by(mtc_county_id, occupation) %>%
  summarise(synthesized_persons = sum(finalweight)) %>%
  mutate(measure = "ERROR") %>%
  mutate(measure = ifelse(occupation == 1, "occupation_management", measure)) %>%
  mutate(measure = ifelse(occupation == 2, "occupation_professional", measure)) %>%
  mutate(measure = ifelse(occupation == 3, "occupation_services", measure)) %>%
  mutate(measure = ifelse(occupation == 4, "occupation_retail", measure)) %>%
  mutate(measure = ifelse(occupation == 5, "occupation_manual", measure)) %>%
  mutate(measure = ifelse(occupation == 6, "occupation_military", measure)) %>%
  group_by(mtc_county_id, measure) %>%
  summarise(synthesized_persons = sum(synthesized_persons)) %>%
  mutate(measure_category_person = "Persons by Occupation")

county_synthesized <- rbind(county_synthesized, working)

county_validation <- left_join(county_validation, county_synthesized, by = c("mtc_county_id", "measure"))

# standardize variables to combine across geographies
county_validation <- county_validation %>%
  rename(geography_index = mtc_county_id) %>%
  select(-county_name) %>%
  mutate(geography = "county") %>%
  mutate(control_households = ifelse(!is.na(synthesized_households), households_or_persons, NA)) %>%
  mutate(control_persons    = ifelse(!is.na(synthesized_persons),    households_or_persons, NA)) %>%
  mutate(measure_category   = ifelse(!is.na(synthesized_persons),    measure_category_person, measure_category_household)) %>%
  mutate(control_category = "controlled") %>%
  select(-households_or_persons, -measure_category_person, -measure_category_household)
           

remove(working, county_synthesized)
              
```

#### County for Uncontrolled Measures
```{r county-uncontrolled}
# build PUMA to county_id crosswalk
puma_county_crosswalk <- maz_controls %>%
  group_by(PUMA, mtc_county_id) %>%
  summarise(count = n()) %>%
  select(-count)

# uncontrolled #1: age by year
working <- pums_person %>%
  select(PUMA, agep = AGEP, PWGTP) %>%
  group_by(PUMA, agep) %>%
  summarise(puma_persons = sum(PWGTP))

working <- left_join(working, puma_county_crosswalk, by = c("PUMA"))

working <- working %>%
  group_by(mtc_county_id, agep) %>%
  summarise(control_persons = sum(puma_persons)) %>%
  ungroup()

county_synthesized <- person_df %>%
  group_by(mtc_county_id, agep) %>%
  summarise(synthesized_persons = sum(finalweight)) %>%
  mutate(measure = paste("age", agep, sep = "_")) %>%
  mutate(measure_category = "Persons by Age") %>%
  ungroup()

working <- left_join(working, county_synthesized, by = c("mtc_county_id", "agep"))

county_cross_validation <- working %>%
  select(-agep)

# uncontrolled #2: race & ethnicity
working <- pums_person %>%
  select(PUMA, rac1p = RAC1P, hisp = HISP, PWGTP) %>%
  group_by(PUMA, rac1p, hisp) %>%
  summarise(puma_persons = sum(PWGTP))

working <- left_join(working, puma_county_crosswalk, by = c("PUMA"))

working <- working %>%
  group_by(mtc_county_id, rac1p, hisp) %>%
  summarise(control_persons = sum(puma_persons)) %>%
  ungroup() 

county_synthesized <- person_df %>%
  group_by(mtc_county_id, rac1p, hisp) %>%
  summarise(synthesized_persons = sum(finalweight)) %>%
  ungroup()

working <- left_join(working, county_synthesized, by = c("mtc_county_id", "rac1p", "hisp"))

working <- working %>%
  mutate(synthesized_persons = ifelse(is.na(synthesized_persons), 0L, synthesized_persons)) %>%
  mutate(measure = "Other") %>%
  mutate(measure = ifelse(hisp > 1, "Hispanic", measure)) %>%
  mutate(measure = ifelse(rac1p == 1 & hisp == 1, "White", measure)) %>%
  mutate(measure = ifelse(rac1p == 2 & hisp == 1, "Black", measure)) %>%
  mutate(measure = ifelse(rac1p == 6 & hisp == 1, "Asian", measure)) %>%
  group_by(mtc_county_id, measure) %>%
  summarize(control_persons = sum(control_persons), synthesized_persons = sum(synthesized_persons)) %>%
  ungroup() %>%
  mutate(measure_category = "Race Ethnicity")

county_cross_validation <- rbind(county_cross_validation, working)

# uncontrolled #3: usual hours worked per week last 12 months
working <- pums_person %>%
  select(PUMA, wkhp = WKHP, PWGTP) %>%
  mutate(wkhp = ifelse(is.na(wkhp), 0L, wkhp)) %>%
  group_by(PUMA, wkhp) %>%
  summarise(puma_persons = sum(PWGTP))

working <- left_join(working, puma_county_crosswalk, by = c("PUMA"))

working <- working %>%
  group_by(mtc_county_id, wkhp) %>%
  summarise(control_persons = sum(puma_persons)) %>%
  ungroup() 

county_synthesized <- person_df %>%
  mutate(wkhp = ifelse(is.na(wkhp), 0L, wkhp)) %>%
  group_by(mtc_county_id, wkhp) %>%
  summarise(synthesized_persons = sum(finalweight)) %>%
  ungroup()

working <- left_join(working, county_synthesized, by = c("mtc_county_id", "wkhp"))

working <- working %>%
  mutate(synthesized_persons = ifelse(is.na(synthesized_persons), 0L, synthesized_persons)) %>%
  mutate(measure = "ERROR") %>%
  mutate(measure = ifelse(wkhp == 0, "Non-worker", measure)) %>%
  mutate(measure = ifelse(wkhp > 0 & wkhp < 21, "Less than 21 hours", measure)) %>%
  mutate(measure = ifelse(wkhp >= 21 &  wkhp < 36, "Between 21 and 35 hours", measure)) %>%
  mutate(measure = ifelse(wkhp >= 36 & wkhp < 46, "Between 35 and 46 hours", measure)) %>%
  mutate(measure = ifelse(wkhp >= 46, "More than 46 hours", measure)) %>%
  group_by(mtc_county_id, measure) %>%
  summarise(control_persons = sum(control_persons), synthesized_persons = sum(synthesized_persons)) %>%
  mutate(measure_category = "Usual Work Week Hours") %>%
  ungroup()

county_cross_validation <- rbind(county_cross_validation, working)

# uncontrolled #4: school grade
working <- pums_person %>%
  select(PUMA, schg = SCHG, PWGTP) %>%
  mutate(schg = ifelse(is.na(schg), 0L, schg)) %>%
  group_by(PUMA, schg) %>%
  summarise(puma_persons = sum(PWGTP))

working <- left_join(working, puma_county_crosswalk, by = c("PUMA"))

working <- working %>%
  group_by(mtc_county_id, schg) %>%
  summarise(control_persons = sum(puma_persons)) %>%
  ungroup() 

county_synthesized <- person_df %>%
  mutate(schg = ifelse(is.na(schg), 0L, schg)) %>%
  group_by(mtc_county_id, schg) %>%
  summarise(synthesized_persons = sum(finalweight)) %>%
  ungroup()

working <- left_join(working, county_synthesized, by = c("mtc_county_id", "schg"))

working <- working %>%
  mutate(synthesized_persons = ifelse(is.na(synthesized_persons), 0L, synthesized_persons)) %>%
  mutate(measure = "ERROR") %>%
  mutate(measure = ifelse(schg == 0, "Non-student", measure)) %>%
  mutate(measure = ifelse(schg == 1, "Preschool", measure)) %>%
  mutate(measure = ifelse(schg == 2 | schg == 3, "Grade School", measure)) %>%
  mutate(measure = ifelse(schg == 4, "Middle School", measure)) %>%
  mutate(measure = ifelse(schg == 5, "High School", measure)) %>%
  mutate(measure = ifelse(schg >= 6, "College", measure)) %>%
  group_by(mtc_county_id, measure) %>%
  summarise(control_persons = sum(control_persons), synthesized_persons = sum(synthesized_persons)) %>%
  mutate(measure_category = "Student Status") %>%
  ungroup()

county_cross_validation <- rbind(county_cross_validation, working)

# add other variables for consistency
county_cross_validation <- county_cross_validation %>%
  rename(geography_index = mtc_county_id) %>%
  mutate(geography = "county") %>%
  mutate(synthesized_households = NA) %>%
  mutate(control_households = NA) %>%
  mutate(control_category = "not_controlled")
  
remove(working, county_synthesized)
```

#### Region for Uncontrolled Measures

#### Bind data across geographies, control level
```{r bind}
# combine controlled and uncontrolled
county_combined_validation <- rbind(county_validation, county_cross_validation)

# combine across geographies
validation_output <- rbind(maz_validation, taz_validation, county_combined_validation)

remove(county_validation, county_cross_validation, maz_validation, taz_validation, county_combined_validation)
```


#### Scale distribution controls/uncontrolled measures
# TODO: is this helpful? should a strong county persons control do the same thing?
```{r scaling}
# table(validation_output$measure_category)
# 
# # compute scaling factors, set to 1.00 for persons and households controls
# scaling_factors <- validation_output %>%
#   group_by(measure_category) %>%
#   summarise(person_adjustment    = sum(synthesized_persons) / sum(control_persons),
#             household_adjustment = sum(synthesized_households) / sum(control_households)) %>%
#   ungroup() %>%
#   mutate(household_adjustment = ifelse(measure_category == "Households", 1.00, household_adjustment)) %>%
#   mutate(person_adjustment    = ifelse(measure_category == "Persons",    1.00, person_adjustment))
# 
# validation_output <- left_join(validation_output, scaling_factors, by = c("measure_category"))
# 
# validation_output <- validation_output %>%
#   mutate(control_households = control_households * household_adjustment) %>%
#   mutate(control_persons    = control_persons * person_adjustment)


```


#### Write to disk
```{r data-writes}

write.csv(validation_output, file = F_OUTPUT, row.names = FALSE, quote = F)
```

