---
title: "Consume Outputs"
author: "David Ory"
output: 
  html_document:
    theme: cosmo
    toc: yes
---


## Administration

#### Purpose
Consumes outputs from the Population Synthesizer. These procedures replace previous procedures done in SQL. A PUMS data dictionary is available [here](http://www2.census.gov/programs-surveys/acs/tech_docs/pums/data_dict/PUMSDataDict13.pdf).   

#### Outputs
1.  TBD

#### TODO
1. Everything, early days

## Procedure

#### Overhead
```{r overhead, results = 'hide'}
library(knitr)
suppressMessages(library(dplyr))
library(stringr)
suppressMessages(library(RMySQL))
suppressMessages(library(hydroGOF))
library(reshape2)
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#### Paramaters
```{r parameters}
MYSQL_SERVER    = "localhost"
MYSQL_DATABASE  = "pop_syn"
MYSQL_USER_NAME = "root" 

YEAR = "year_2010"

HH_INCOME_QUARTILE_THRESH_1 = 30000
HH_INCOME_QUARTILE_THRESH_2 = 60000
HH_INCOME_QUARTILE_THRESH_3 = 100000
```

#### Remote file locations
```{r remote-dir}
MYSQL_PASSWORD_FILE <- "C:/Users/dory/Desktop/mysql.csv"

```

#### Data reads
```{r data-reads}
# get password
mysql_passes <- read.csv(MYSQL_PASSWORD_FILE, header = TRUE)

mysql_passes <- mysql_passes %>%
  filter(user == MYSQL_USER_NAME) %>%
  mutate(pwd = paste(pwd))

# connection
mysql_connection <- dbConnect(MySQL(), user = MYSQL_USER_NAME, password = mysql_passes$pwd, host = MYSQL_SERVER, dbname = MYSQL_DATABASE)

# read the hh table
household_df <- dbReadTable(conn = mysql_connection, name = 'synpop_hh')
person_df <- dbReadTable(conn = mysql_connection, name = 'synpop_person')

```

#### Validation exercise #1: how closely do we match the controls?
```{r validate-controls}
maz_controls <-    dbReadTable(conn = mysql_connection, name = 'control_totals_maz')
taz_controls <-    dbReadTable(conn = mysql_connection, name = 'control_totals_taz')
county_controls <- dbReadTable(conn = mysql_connection, name = 'control_totals_meta')

dbDisconnect(mysql_connection)

# step 1: maz controls
maz_synthezised <- household_df %>%
  group_by(maz) %>%
  summarise(synthesized_households = sum(finalweight))

maz_validation <- maz_controls %>%
  select(maz, control_households = households)

maz_validation <- left_join(maz_validation, maz_synthezised, by = c("maz"))

maz_validation <- maz_validation %>%
  mutate(synthesized_households = ifelse(is.na(synthesized_households), 0L, synthesized_households))

maz_rmse <- rmse(maz_validation$synthesized_households, maz_validation$control_households, na.rm = FALSE)

# put results in validation data frame
control_name <- c('Households')
geography <- c('MAZ')
count <- nrow(maz_validation)
control_quantity <- sum(maz_validation$control_households)
synthesized_quantity <- sum(maz_validation$synthesized_households)
root_mean_square_error <- maz_rmse
validation_df <- data.frame(control_name, geography, count, control_quantity, synthesized_quantity, root_mean_square_error) 
  
# START HERE: make sure this is working
# step 2: taz controls
taz_synthesized <- household_df %>%
  mutate(income_quartile = 0L) %>%
  mutate(income_quartile = ifelse(hh_income_2010 < HH_INCOME_QUARTILE_THRESH_1, 1L, income_quartile)) %>%
  mutate(income_quartile = ifelse(hh_income_2010 >= HH_INCOME_QUARTILE_THRESH_1 & hh_income_2010 < HH_INCOME_QUARTILE_THRESH_2, 2L, income_quartile)) %>%
  mutate(income_quartile = ifelse(hh_income_2010 >= HH_INCOME_QUARTILE_THRESH_2 & hh_income_2010 < HH_INCOME_QUARTILE_THRESH_3, 3L, income_quartile)) %>%
  mutate(income_quartile = ifelse(hh_income_2010 >= HH_INCOME_QUARTILE_THRESH_3, 4L, income_quartile)) %>%
  group_by(taz, income_quartile) %>%
  summarise(synthesized_households = sum(finalweight))

taz_validation <- taz_controls %>%
  select(taz, one = hh_income_quartile_1, two = hh_income_quartile_2, three = hh_income_quartile_3, four = hh_income_quartile_4)

taz_validation <- melt(taz_validation, id = c("taz"), variable.name = "income_quartile", value.name = "control_households")

taz_validation <- taz_validation %>%
  mutate(income_quartile = ifelse(income_quartile == "one",   1L, income_quartile)) %>%
  mutate(income_quartile = ifelse(income_quartile == "two",   2L, income_quartile)) %>%
  mutate(income_quartile = ifelse(income_quartile == "three", 3L, income_quartile)) %>%
  mutate(income_quartile = ifelse(income_quartile == "four",  4L, income_quartile))

taz_validation <- left_join(taz_validation, taz_synthesized, by = c("taz", "income_quartile"))

taz_validation <- taz_validation %>%
  mutate(synthesized_households = ifelse(is.na(synthesized_households), 0L, synthesized_households)) %>%
  mutate(control_households     = ifelse(is.na(control_households),     0L, control_households))

# one entry per hh income quartile
for (working_income_quartile in 1:4) {
  
  working_df <- taz_validation %>%
    filter(income_quartile == working_income_quartile)
  
  append_df <- data.frame(control_name = c(paste('Households in Income Quartile',working_income_quartile, sep = ' ')),
                          geography = c('TAZ'),
                          count = nrow(working_df),
                          control_quantity = sum(working_df$control_households),
                          synthesized_quantity = sum(working_df$synthesized_households),
                          root_mean_square_error = rmse(as.numeric(working_df$synthesized_households), as.numeric(working_df$control_households), na.rm = FALSE))
  
  validation_df <- rbind(validation_df, append_df)
}



```

#### Validation exercise #2: how closely do we match variables we did not control for (e.g., person type, family structures)?
#### Validation exercise #3: how well does the scheme perform in a backcast?



